An empirical assessment of strategies to model opponent effects in crash severity analysis

### 2020-01-15 Interactions-Crash-Severity-v1/Interactions-Crash-Severity-v1.tex
Added Model 4 ensemble and verified all models for convergence

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={An assessment of strategies to model opponent effects in crash severity analysis},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)
% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\usepackage[margin=1in]{geometry}

@@

  \title{An assessment of strategies to model opponent effects in crash severity
analysis}
    \author[Some University]{Author 1\corref{c1}}
   \ead{a1@example.com} 
   \cortext[c1]{Corresponding Author}
    \author[Some Institute of Technology]{Author 2}
   \ead{a2@example.com}

      \address[Some University]{Department, Street, City, State, Zip}
    \address[Some Institute of Technology]{Department, Street, City, State, Zip}

@@

The rest of this paper is structured as follows. In Section
\ref{sec:review-of-methods} we present a concise review of the methods
used to analyze crash severity, with a particular focus on techniques
that consider the interactions between participants in a crash. Section
\ref{sec:methods} describes the data requirements, data preprocessing,
and the modelling strategy. Model estimation is presented in
\ref{sec:application} and the assessment of models is in Section
\ref{sec:assessment}. We then present some additional thoughts about the
applicability of this approach in Section
\ref{sec:further-considerations} before offering some concluding remarks
in Section \ref{sec:concluding-remarks}.

@@

@@

Crash severity is often modelled using models for discrete outcomes,
including classification techniques from machine learning (e.g.,
Iranitalab and Khattak, 2017; Chang and Wang, 2006; Effati et al., 2015;
Khan et al., 2015), Poisson models for counts (e.g., Ma et al., 2008),
unordered logit/probit models {[}{]}, as well as ordered logit/probit
models (e.g., Rifaat and Chin, 2007), with numerous variants, such as
random parameters/mixed logit (e.g., Aziz et al., 2013; Haleem and Gan,
2013), partial proportional odds models (e.g., Mooradian et al., 2013;
Sasidharan and Menendez, 2014), and the use of copulas (e.g., Wang et
al., 2015). Recent reviews of methods include Savolainen et al.~(2011)
and Shamsunnahar and Eluru (2013). Shamsunnahar and Eluru (2013) in
particular conducted an extensive comparison of models for discrete
outcomes and found that while the difference between the performance of
unordered models and ordered models was so small as to make not
difference, ordered models are usually more parsimonious since only one
latent functions needs to be estimated for all outcomes, as opposed to
one for each outcome in unordered modelling mechanisms.

Irrespective of the modelling framework employed, models of crash
severity often include variables in several categories, as shown with
examples in Table \ref{tab:table-variable-categories} (also see Montella
et al., 2013). Many crash databases (but not by any means all) also
account for the multievent nature of many crashes. In this way, there
are crashes that involve a single traffic unit (e.g., Kim et al., 2013;
Gong and Fan, 2017), others that involve two traffic units (e.g., Tarrao
et al., 2014; Wang et al., 2015), and more rarely there are
multi-traffic unit crashes (e.g., Wu et al., 2014; Bogue et al., 2017).
Likewise, each traffic unit can possibly involve more than one person
(e.g., driver and passengers). Thus, depending on the incident, each
record in a crash database may include unique identifier for the crash,
as well as identifiers for the traffic units (or identifiers for dummy
objects such as a light pole that was hit by a vehicle), and the people
involved in the crash.

@@

\caption{\label{tab:table-variable-categories}\label{tab:table-variable-categories}Categories of variables used in the analysis of crash severity with examples}
\centering
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{22em}}
\toprule
Category & Examples\\
\midrule
Person-related & Attributes of participants in the crash, e.g., injury status, age, gender, licensing status, professional driver status\\
Traffic unit-related & Attributes of the traffic unit, e.g., type of traffic unit (car, motorcycle, etc.), maneouver, etc.\\
Crash-related & Attributes of the crash, e.g., location, weather conditions, light conditions, number of parties, etc.\\
Road-related & Attributes of the road, e.g., surface condition, grade, geometry, etc.\\
\bottomrule
\end{tabular}
\end{table}

Interplay between participants:

Chiou et al. (2013)

Lee and Li (2014) consider effect on crash severity of interactions
between different types of vehicles. This they do by subsetting the
dataset and estimating independent models for each subset of data. Since
they consider three types of vehicles, namely cars (C), light trucks
(L), and heavy trucks (H), they work with nine datasets, for each type
of interactions (i.e., C-C, C-L, C-H, and so on). Tarrao et al. (2014)

Mannering et al. (2016)

Li et al. (2017)

Salon and McIntyre (2018)

Chen et al. (2019)

Wu et al. (2014) two-vehicle crashes do not really consider the
interactions.

Many models use a latent-variable approach, whereby the severity of the
crash (observed) is linked to an underlying latent variable that is a
function of the variables: \[
y_{ik}^*=\sum_{l=1}\beta_lp_{ikl} + \sum_{m=1}\gamma_mc_{km} + \epsilon_{ik}
\]

The left-hand side of the expression above (\(y_{ik}^*\)) is a latent
(unobservable) variable that is associated with the severity of crash
\(k\) (\(k=1,\cdots,K\)) for participant \(i\) (\(i = 1,\cdots,n\)). The
right-hand side of the expression is split in three parts. The first
part collects \(l=1,\cdots,L\) individual attributes \(p\) for
participant \(i\) in crash \(k\); these could relate to the person
(e.g., age and gender) or be individual attributes of the traffic unit
(e.g., maneuver or vehicle type). The second part collects
\(m=1,\cdots,M\) attributes \(c\) related to the crash \(k\), including
crash-related and road-related data. The last part is a random term
specific to participant \(i\) in crash \(k\).

The latent variable is not observed directly, but it is possible to
posit a probabilistic relationship with the outcome \(y_{ik}\) (the
severity of crash \(k\) for participant \(i\)). Depending on the
characteristics of the data and the assumptions made about the random
component of the latent function different models can be obtained. For
example, if crash severity is coded as a binary variable (e.g.,
non-fatal/fatal), we can relate the latent variable to the outcome as
follows: \[
y_{ik} = 
\begin{cases}
\text{fatal} & \text{if } y_{ik}^*>0\\
\text{non-fatal} & \text{if } y_{ik}^*\leq0
\end{cases}
\]

Due to the stochastic nature of the latent function, the outcome of the
crash is not fully determined. However, we can make the following
probability statement: \[
P(y_i = \text{fatal}) = P(y_i^* > 0)
\] In other words, the probability of a fatal accident equals the
probability that the latent variable is greater than zero. This implies
(see Maddala, 1986, p. 22): \[
\begin{array}{rl}\
P(y_i = \text{fatal}) &= P(\sum_{l=1}\beta_lp_{ikl} + \sum_{m=1}\gamma_mc_{km} + \epsilon_{ik} > 0)\\ 
&=P(\epsilon_{ik} > -\sum_{l=1}\beta_lp_{ikl} - \sum_{m=1}\gamma_mc_{km})
\end{array}
\]

It the random terms \(\epsilon_i\) are assumed to follow the logistic
distribution, the the binary logit model is obtained; if they are
assumed to follow the normal distribution, the binary probit model is
obtained.

More often, though, the outcome is recorded using more categories, for
example property damage only (PDO)/injury/fatality. A similar approach
can be adopted, with a latent variable that relates to the outcome as
follows: \[
y_i = 
\begin{cases}
\text{fatality} & \text{if } y_i^*> k_2\\
\text{injury} & \text{if } k_1< y_i^*<k_2\\
\text{PDO} & \text{if } y_i^*<k_1
\end{cases}
\] where \(k_1\) and \(k_2\) are estimable thresholds. In this case, the
associated probability statements are as follows: \[
\begin{array}{rcl}\
P(y_{ik} = \text{PDO}) &=& 1 - P(y_{ik} = \text{injury}) - P(y_{ik} = \text{fatality})\\ 
P(y_{ik} = \text{injury}) &=& P(k_1 - \sum_{l=1}\beta_lp_{ikl} - \sum_{m=1}\gamma_mc_{km} < \epsilon_{ik} < k_2 - \sum_{l=1}\beta_lp_{ikl} - \sum_{m=1}\gamma_mc_{km} )\\
P(y_{ik} = \text{fatality}) &=& P(\epsilon_{ik} < k_1 - \sum_{l=1}\beta_lp_{ikl} - \sum_{m=1}\gamma_mc_{km} )
\end{array}
\]

@@

\hypertarget{sec:methods}{%
\section{Methods}\label{sec:methods}}

\hypertarget{data-considerations}{%
\subsection{Data considerations}\label{data-considerations}}

Words go here

Montella et al. (2013)

\hypertarget{model-specification}{%
\subsection{Model specification}\label{model-specification}}

More words go here

\hypertarget{sec:application}{%
\section{Application}\label{sec:application}}

\begin{quote}
\textbf{Note:} that this paper presents reproducible research. The
source file is an R Markdown document. All code and data necessary to
reproduce the analysis are available from the following anonymous Drive
folder:
\end{quote}

@@

\hypertarget{data-for-the-application}{%
\subsection{Data for the application}\label{data-for-the-application}}

To assess the performance of the modelling strategies discussed in
Section \ref{sec:methods}, we use data from Canada's National Collision
Database (NCDB). This is database contains all motor vehicle collisions
on public roads in Canada as reported by a police service. Data are
collected by provinces and territories, and shared with the federal
government, where data are combined, tracked, and analyzed for reporting
of deaths, injuries, and collisions in Canada at the national level. The
NCDB is provided by Transport Canada, the agency of the federal
government of Canada in charge of transportation policies and programs,
under the Open Government License - Canada version 2.0
{[}https://open.canada.ca/en/open-government-licence-canada{]}.

The NCDB is available from 1999. For the purpose of this paper, we use
the most recent year available as of this writing (2017). Furthermore,
for assessment we also use the data corresponding to 2016. Similar to
databases in other jurisdictions {[}see @{]}, the NCDB contains
information pertaining to the collision, the traffic unit, and the
person(s) involved.

\begin{tabular}{ll}
\toprule
Variable & Details\\
\midrule
Age & Details\\
\bottomrule
\multicolumn{2}{l}{\textit{Note: }}\\
\multicolumn{2}{l}{NCDB available from https://open.canada.ca/data/en/dataset/1eb9eba7-71d1-4b30-9fb1-30cbdab7e63a}\\
\end{tabular}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={An empirical assessment of strategies to model opponent effects in crash severity analysis},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc header
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

  \title{An empirical assessment of strategies to model opponent effects in crash
severity analysis}
    \author[Some University]{Author 1\corref{Corresponding Author}}
   \ead{a1@example.com} 
    \author[Some Institute of Technology]{Author 2}
   \ead{a2@example.com} 
      \address[Some University]{Department, Street, City, State, Zip}
    \address[Some Institute of Technology]{Department, Street, City, State, Zip}

@@

The rest of this paper is structured as follows. In Section
\ref{sec:review-of-methods} we present a concise review of the methods
used to analyze crash severity, with a particular focus on techniques
that consider the interactions between participants in a crash. Section
\ref{sec:application} describes the data requirements, data
preprocessing, and the modelling strategy, along with the results of
model estimation. The results of assessing the models and the discussion
of these results is found in Section \ref{sec:assessment}. We then
present some additional thoughts about the applicability of this
approach in Section \ref{sec:further-considerations} before offering
some concluding remarks in Section \ref{sec:concluding-remarks}.

@@

\hypertarget{general-considerations}{%
\subsection{General considerations}\label{general-considerations}}

@@

Crash severity is often modelled using models for discrete outcomes,
including classification techniques from machine learning (e.g.,
Iranitalab and Khattak, 2017; Chang and Wang, 2006; Effati et al., 2015;
Khan et al., 2015), Poisson models for counts (e.g., Ma et al., 2008),
unordered logit/probit models (e.g., Tay et al., 2011), as well as
ordered logit/probit models (e.g., Rifaat and Chin, 2007), with numerous
variants, such as random parameters/mixed logit (e.g., Aziz et al.,
2013; Haleem and Gan, 2013), partial proportional odds models (e.g.,
Mooradian et al., 2013; Sasidharan and Menendez, 2014), and the use of
copulas (e.g., Wang et al., 2015). Recent reviews of methods include
Savolainen et al.~(2011) and Shamsunnahar and Eluru (2013).

Irrespective of the modelling framework employed, models of crash
severity often include variables in several categories, as shown with
examples in Table \ref{tab:variable-categories} (also see Montella et
al., 2013). Many crash databases and analyses also account for the
multievent nature of many crashes. Participants may have had different
roles in a crash depending on their context, with some acting as
operators of a vehicle (i.e., drivers, bicyclists), while others were
passengers. They also may differ depending on what type of traffic unit
they were, for example occupants of a light duty vehicle or a truck,
motorcyclists, or pedestrians. The multiplicity of roles makes for
complicated modelling choices when trying to understand the severity of
injuries; for example, what is the unit of analysis, the person, the
traffic unit, or the collision? Not surprisingly, it is possible to find
examples of studies that adopt different perspectives. A common
simplifying strategy to specify a model is to consider only
\emph{drivers} and/or only \emph{single-vehicle} crashes (e.g., Kim et
al., 2013; Gong and Fan, 2017; Lee and Li, 2014; Osman et al., 2018).
This strategy reduces the dimensions of the event, and it becomes
possible, for example, to equate the traffic unit to the person for
modelling purposes.

The situation becomes more complex when dealing with events that involve
two traffic units (e.g., Tarrao et al., 2014; Wang et al., 2015) and
multi-traffic unit crashes (e.g., Wu et al., 2014; Bogue et al., 2017).
Different strategies have been developed to study these, more complex
cases. A number of studies advocate the estimation of separate models
for different participants and/or situations. In this way, Wang and
Kockelman (2005) estimated models for single-vehicle and two-vehicle
crashes, while Savolainen and Mannering (2007) estimated models for
single-vehicle and multi-vehicle crashes. More recently, Duddu et
al.~(2018) and Penmetsa et al.~(2017) presented research that estimated
separate models for at-fault and not-at-fault drivers. The strategy of
estimating separate models also relies on subsetting the dataset,
although it is possible to link the relevant models more tightly by
means of a common covariance structure, as is the case of bivariate
models (e.g., Chiou et al., 2013; Chen et al., 2019) or models with
copulas (e.g., Rana et al., 2010; Shamsunnahar et al., 2014; Wang et
al., 2015).

A related strategy to specify a crash severity model is to organize the
data in such a way that it is possible to model the influence of the
attributes of the opponent in a crash. There are numerous examples of
studies that consider at least some characteristics of the oponents in
two- or multi-vehicle crashes. For example, Wang and Kockelman (2005)
considered the type of the opposing vehicle in their model for
two-vehicle collisions. Similarly, Tarrao et al.~(2014) included in
their analysis the age, wheelbase, weight, and engine size of the
opposing vehicle while Bogue et al.~(2017) used the body type of the
opposing vehicle. Penmetsa et al.~(2017) and Duddu et al.~(2018) are two
of the most comprehensive examples of using opponent's information, with
the physical condition, sex, age, and vehicle type of the opponent. The
twin strategies of subsetting the sample and using the attributes of the
opponent are not mutually exclusive, but neither are they consistently
used together, as a scan of the literature reveals.

@@

\caption{\label{tab:table-variable-categories}\label{tab:variable-categories}Categories of variables used in the analysis of crash severity with examples}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{22em}}
\toprule
Category & Examples\\
\midrule
Person-related & Attributes of participants in the crash, e.g., injury status, age, gender, licensing status, professional driver status\\
Traffic unit-related & Attributes of the traffic unit, e.g., type of traffic unit (car, motorcycle, etc.), maneouver, etc.\\
Crash-related & Attributes of the crash, e.g., location, weather conditions, light conditions, number of parties, etc.\\
Road-related & Attributes of the road, e.g., surface condition, grade, geometry, etc.\\
Opponent-related & Attributes of the opponent, e.g., age of opponent, gender of opponent, opponent vehicle type, etc.\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{modelling-techniques}{%
\subsection{Modelling techniques}\label{modelling-techniques}}

With respect to model structures, Shamsunnahar and Eluru (2013)
conducted an extensive comparison of models for discrete outcomes and
found only small differences in the performance of unordered models and
ordered models; however, ordered models are usually more parsimonious
since only one latent functions needs to be estimated for all outcomes,
as opposed to one for each outcome in unordered modelling mechanisms.
Bogue et al.~(2017) also compared unordered and ordered models in the
form of the mixed multinomial logit and a modified rank ordered logit,
respectively, and found that the ordered model performed best. To keep
the empirical assessment managable we will consider only the ordinal
logit model, and will comment on potential extensions in Section
\ref{sec:further-considerations}.

The ordinal model is a latent-variable approach, whereby the severity of
the crash (observed) is linked to an underlying latent variable that is
a function of the variables of interest, as follows:

\begin{equation}
\label{eq:latent-function}
y_{itk}^*=\sum_{l=1}^L\alpha_lp_{itkl} + \sum_{m=1}^M\beta_mu_{tkm} + \sum_{q=1}^Q\kappa_qc_{kq} + \epsilon_{itk}
\end{equation}

The left-hand side of the expression above (\(y_{itk}^*\)) is a latent
(unobservable) variable that is associated with the severity of crash
\(k\) (\(k=1,\cdots,K\)) for participant \(i\) in traffic unit \(t\).
The right-hand side of the expression is split in four parts. The first
part collects \(l=1,\cdots,L\) individual attributes \(p\) for
participant \(i\) in traffic unit \(t\) and crash \(k\); these could
relate to the person (e.g., age, gender, and road user class). The
second part collects \(m=1,\cdots,M\) attributes \(u\) related to
traffic unit \(t\) in crash \(k\); these could be items such as maneuver
or vehicle type. The third part collects \(q=1,\cdots,Q\) attributes
\(c\) related to the crash \(k\), including crash-related and
road-related data, such as weather conditions, road alignment, and type
of surface. Lastly, the fourth element is a random term specific to
participant \(i\) in traffic unit \(t\) and crash \(k\). The function
consists of a total of \(Z=L+M+Q\) covariates and associated parameters.

When opponent-related variables are included, the function is augmented
as follows:

\begin{equation}
\label{eq:latent-function-with-opponent-variables}
y_{itk}^*=\sum_{l=1}^L\alpha_lp_{itkl} + \sum_{m=1}^M\beta_mu_{tkm} + \sum_{q=1}^Q\kappa_qc_{kq} + \sum_{r=1}^R\delta_ro_{jvkr} + \epsilon_{itk}
\end{equation}

The additional part collects \(r=1,\cdots,R\) attributes \(o\) related
to individual \(j\) in traffic unit \(v\) and crash \(k\) that opposed
individual \(i\) in traffic unit \(t\) and crash \(k\). These could be
individual characteristics of the opponent (such as age and gender)
and/or characteristics of the opposing vehicle (such as vehicle type or
weight). To qualify as an opponent, individual \(j\) must have been a
participant in crash \(k\) but operating traffic unit \(v\ne t\).
Sometimes the person \emph{is} the traffic unit, as is the case of a
pedestrian. And we exclude passengers of vehicles as opponents, since
they do not operate the traffic unit. In case the opponent attributes
include only characteristics of the traffic unit, the condition for the
traffic unit to be an opponent is that it participated in crash \(k\)
and was different from \(t\). After introducing this new set of terms,
the latent function now consists of a total of \(Z=L+M+Q+R\) covariates
and associated parameters.

For conciseness, in what follows we will abbreviate the function as
follows:

\begin{equation}
\label{eq:latent-function-compact}
y_{itk}^*=\sum_{z=1}^Z\theta_zx_{itkz} + \epsilon_{itk}
\end{equation}

The latent variable is not observed directly, but it is possible to
posit a probabilistic relationship with the outcome \(y_{ik}\) (the
severity of crash \(k\) for participant \(i\)). Depending on the
characteristics of the data and the assumptions made about the random
component of the latent function different models can be obtained. For
example, if crash severity is coded as a binary variable (e.g.,
non-fatal/fatal), we can relate the latent variable to the outcome as
follows:

\begin{equation}
\label{eq:latent-function-2-outcomes}
y_{itk} = 
\begin{cases}
\text{fatal} & \text{if } y_{itk}^*>0\\
\text{non-fatal} & \text{if } y_{itk}^*\leq0
\end{cases}
\end{equation}

Due to the stochastic nature of the latent function, the outcome of the
crash is not fully determined. However, we can make the following
probability statement:

\begin{equation}
\label{eq:probability-2-outcomes}
P(y_{itk} = \text{fatal}) = P(y_{itk}^* > 0)
\end{equation}

In other words, the probability that individual \(i\) in traffic unit
\(k\) and crash \(k\) was a fatality equals the probability that the
latent variable is greater than zero. This implies (see Maddala, 1986,
p. 22):

\begin{equation}
\label{eq:probability-2-outcomes-2}
\begin{array}{rl}\
P(y_{itk} = \text{fatal}) &= P(\sum_{z=1}\theta_zp_{itkz} + \epsilon_{itk} > 0)\\ 
&=P(\epsilon_{itk} > -\sum_{z=1}\theta_zp_{itkz})
\end{array}
\end{equation}

If the random terms \(\epsilon_{itk}\) are assumed to follow the
logistic distribution, the the binary logit model is obtained; if they
are assumed to follow the normal distribution, the binary probit model
is obtained. More often, though, the outcome is recorded using more
categories, for example property damage only (PDO)/injury/fatality. A
similar approach can be adopted, with a latent variable that relates to
the outcome as follows:

\begin{equation}
\label{eq:latent-function-ordered-outcomes}
y_{itk} = 
\begin{cases}
\text{fatality} & \text{if } y_{itk}^*> \mu_2\\
\text{injury} & \text{if } \mu_1< y_{itk}^*< \mu_2\\
\text{PDO} & \text{if } y_{itk}^*< \mu_1
\end{cases}
\end{equation}

\noindent where \(\mu_1\) and \(\mu_2\) are estimable thresholds. In
this case, the associated probability statements are as follows:

\begin{equation}
\label{eq:probability-ordered-outcomes}
\begin{array}{rcl}\
P(y_{itk} = \text{PDO}) &=& 1 - P(y_{itk} = \text{injury}) - P(y_{itk} = \text{fatality})\\ 
P(y_{itk} = \text{injury}) &=& P(\mu_1 - \sum_{z=1}\theta_zp_{itkz} < \epsilon_{itk} < \mu_2 - \sum_{z=1}\theta_zp_{itkz})\\
P(y_{itk} = \text{fatality}) &=& P(\epsilon_{itk} < \mu_1 - \sum_{z=1}\theta_zp_{itkz})
\end{array}
\end{equation}

@@

\hypertarget{model-specification-strategies}{%
\subsection{Model specification
strategies}\label{model-specification-strategies}}

In this paper we consider three general model specification strategies,
as follows:

\begin{itemize}
\tightlist
\item
  Strategy 1. Introducing opponent-related variables
\item
  Strategy 2. Single-level model and multi-level (hierarchical) model
  specifications
\item
  Strategy 3. Full sample and sample subsetting
\end{itemize}

Introduction of opponent related-variables was explained in the
preceding subsection. In this way, a base model is given by Equation
\ref{eq:latent-function}, whereas Strategy 1 (inclusion of
opponent-related variables) is Equation
\ref{eq:latent-function-with-opponent-variables}. These two equations
are also examples of single-level model. Next we describe Strategies 2
and 3.

\hypertarget{hierarchical-model-specification}{%
\subsubsection{Hierarchical model
specification}\label{hierarchical-model-specification}}

We can choose to conceptualize the event leading to the outcome as a
hierarchical process. There are a few different ways of doing this. For
example, the hierarchy could be based on individuals in traffic units.
In this case, we can rewrite the latent function as follows:

\begin{equation}
\label{eq:latent-function-with-hierarchical-traffic-unit}
y_{itk}^*=\sum_{m=1}^M\beta_mu_{tkm} + \sum_{q=1}^Q\kappa_qc_{kq} + \sum_{r=1}^R\delta_ro_{jvkr} + \epsilon_{itk}
\end{equation}

The coefficients of the traffic unit nest the individual attributes as
follows. For any given coefficient \(q\):

\begin{equation}
\label{eq:hierarchical-traffic-unit-coefficients}
\beta_{m}=\sum_{l=1}^L\beta_{ml}p_{itkl} 
\end{equation}

Therefore, the corresponding term in the latent function becomes
(assuming that \(p_{itk1} = 1\), i.e., it is a constant term):

\begin{equation}
\label{eq:hierarchical-traffic-unit-coefficients}
\begin{array}{rcl}\
\beta_{m}u_{tkm} &=& \big( \beta_{m1} + \beta_{m2}p_{itk2} + \cdots + \beta_{mL}p_{itkL}\big)u_{tkm}\\ 
 &=& \beta_{m1}u_{tkm} + \beta_{m2}p_{itk2}u_{tkm} + \cdots + \beta_{mL}p_{itkL}u_{tkm}
\end{array}
\end{equation}

As an alternative, the nesting unit could be the interaction
person-opponent, in which case the opponent-level attributes are nested

\begin{equation}
\label{eq:latent-function-with-opponent-variables}
y_{itk}^*=\sum_{l=1}^L\alpha_lp_{itkl} + \sum_{m=1}^M\beta_mu_{tkm} + \sum_{q=1}^Q\kappa_qc_{kq} + \epsilon_{itk}
\end{equation}

\noindent with any person-level coefficient \(l\) that we wish to expand
defined as follows:

\begin{equation}
\label{eq:hierarchical-traffic-unit-coefficients}
\alpha_{l}=\sum_{r=1}^R\alpha_{lr}o_{jvkr}
\end{equation}

\noindent with the same conditions as before, that \(j\ne i\) is the
operator of traffic unit \(v\ne t\). The corresponding term in the
latent function is now (assuming that \(o_{jvk1}=1\), i.e., it is a
constant term):

\begin{equation}
\label{eq:hierarchical-traffic-unit-coefficients}
\begin{array}{rcl}\
\alpha_{l}p_{itkl} &=& \big(\alpha_{l1} + \alpha_{l2}o_{jvk2} + \cdots + \alpha_{lR}o_{jvkR} \big)p_{itkl}\\
 &=& \alpha_{l1}p_{itkl} + \alpha_{l2}o_{jvk2}p_{itkl} + \cdots + \alpha_{lR}o_{jvkR}p_{itkl}
\end{array} 
\end{equation}

Alterted readers will identify this model specification strategy as
Casetti's expansion method (Casetti, 1972). This is a deterministic
strategy for modelling contextual effects which, when augmented with
random components, becomes the well-known multi-level modelling method
(Hedeker and Gibbons, 1994, more on this in Section
\ref{sec:further-considerations}). It is worthwhile to note that
higher-order hierarchical effects are possible; for instance, individual
attributes nested within traffic units, which in turn are nested within
collisions. We do not explore this further in the current paper.

\hypertarget{sample-subsetting}{%
\subsubsection{Sample subsetting}\label{sample-subsetting}}

The third model specification strategy that we will consider is
subsetting the sample. This is applicable in conjunction with any of the
other strategies discussed above. In essence, we define the latent
function with restrictions as follows. Consider a continuous variable,
e.g., age of person, and we wish to concentrate the analysis on older
adults (e.g., Dissanayake and Lu, 2002). The latent function is defined
as desired (see above), however, the following restriction might be
applied to the sample

\begin{equation}
\label{eq:sampling-age}
\text{Age of individual } i \text{ in traffic unit } t \text{ in crash } k = 
\begin{cases}
\ge 65 & \text{use record } itk\\
< 65 & \text{do not use record } itk
\end{cases}
\end{equation}

Suppose instead that we are interested in crashes by or against a
specific type of traffic unit (e.g., pedestrians, Amoh-Gyimah et al.,
2017):

\begin{equation}
\label{eq:sampling-pedestrian}
\text{Road user class of individual } i \text{ in traffic unit } t \text{ in crash } k = 
\begin{cases}
\text{Pedestrian} & \text{use record } itk\\
\text{Not pedestrian} & \text{do not use record } itk
\end{cases}
\end{equation}

\noindent or:

\begin{equation}
\label{eq:sampling-pedestrian-opponent}
\text{Road user class of individual } j \text{ in traffic unit } v \text{ in crash } k = 
\begin{cases}
\text{Pedestrian} & \text{use record } jvk\\
\text{Not pedestrian} & \text{do not use record } jvk
\end{cases}
\end{equation}

More generally, for any variable \(x\) of interest:

\begin{equation}
\label{eq:sampling-general}
x_{itk} = 
\begin{cases}
\text{Condition: TRUE} & \text{use record } itk\\
\text{Condition: FALSE} & \text{do not use record } itk
\end{cases}
\end{equation}

Several conditions can be imposed to subset the sample in any way that
the analyst deems appropriate or suitable.

\hypertarget{sec:application}{%
\section{Setting for empirical assessment}\label{sec:application}}

In this section we present the setting for the empirical assessment of
the modelling strategies discussed in Section
\ref{sec:review-of-methods}, namely matters related to data and model
estimation.

\begin{quote}
\textbf{Note:} this paper presents reproducible research. The source
file is an R Markdown document. All code and data necessary to reproduce
the analysis are available from the following anonymous Drive folder:
\end{quote}

@@

\hypertarget{data-for-empirical-assessment}{%
\subsection{Data for empirical
assessment}\label{data-for-empirical-assessment}}

To assess the performance of the various modelling strategies we use
data from Canada's National Collision Database (NCDB). This is database
contains all motor vehicle collisions on public roads in Canada as
reported by a police service. Data are collected by provinces and
territories, and shared with the federal government, where data are
combined, tracked, and analyzed for reporting of deaths, injuries, and
collisions in Canada at the national level. The NCDB is provided by
Transport Canada, the agency of the federal government of Canada in
charge of transportation policies and programs, under the Open
Government License - Canada version 2.0
{[}https://open.canada.ca/en/open-government-licence-canada{]}.

The NCDB is available from 1999. For the purpose of this paper, we use
the data corresponding to 2017, which is the most recent year available
as of this writing. Furthermore, for assessment we also use the data
corresponding to 2016. Similar to databases in other jurisdictions (see
Montella et al., 2013), the NCDB contains information pertaining to the
collision, the traffic unit(s), and the person(s) involved in a crash,
as shown in Tables \ref{tab:ncdb-descriptives-collision},
\ref{tab:ncdb-descriptives-vehicle}, and
\ref{tab:ncdb-descriptives-person}. Notice that, compared to Table
\ref{tab:variable-categories}, crash-related variables and road-related
variables are collected in a single variable class, namely
collision-related, since they are unique for each crash.

Data are organized by person; in other words, there is one record per
participant in a collision, be they drivers, passengers, pedestrians,
etc. The only variable directly available with respect to opponents in a
collision is the number of vehicles involved (see models in Bogue et
al., 2017). Therefore, the data needs to be processed to obtain
attributes of opponents for each participant in a collision. The
protocol to do this is described next.

\begingroup\fontsize{7}{9}\selectfont

\begin{longtable}[t]{ll>{\raggedright\arraybackslash}p{32em}}
\caption{\label{tab:ncdb-descriptives-collision}\label{tab:ncdb-descriptives-collision}Contents of National Collision Database: Collision-level variables}\\
Variable & Description & Notes\\
\endfirsthead
\caption[]{\label{tab:ncdb-descriptives-collision}Contents of National Collision Database: Collision-level variables \textit{(continued)}}\\
Variable & Description & Notes\\
\endhead
\
\endfoot
\multicolumn{3}{l}{\textit{Note: }}\\
\multicolumn{3}{l}{Source NCDB available from https://open.canada.ca/data/en/dataset/1eb9eba7-71d1-4b30-9fb1-30cbdab7e63a}\\
\multicolumn{3}{l}{Preprocessed data for analysis available from https://drive.google.com/open?id=12aJtVBaQ4Zj0xa7mtfqxh0E48hKCb\_XV}\\
\endlastfoot
\rowcolor{gray!6}  C\_CASE & Unique collision identifier & Unique identifier for collisions\\
C\_YEAR & Year & Last two digits of year.\\
\rowcolor{gray!6}  C\_MNTH & Month & 14 levels: January - December; unknown; not reported by jurisdiction.\\
C\_WDAY & Day of week & 9 levels: Monday - Sunday; unknown; not reported by jurisdiction.\\
\rowcolor{gray!6}  C\_HOUR & Collision hour & 25 levels: hourly intervals; unknown; not reported by jurisdiction.\\
\addlinespace
C\_SEV & Collision severity & 4 levels: collision producing at least one fatality; collision producing non-fatal injury; unknown; not reported by jurisdiction.\\
